---
title: "Exploration of the events.csv dataset"
author: "Daniel Da Graça"
date: "27/05/2023"
output: 
  html_document: 
    output_dir: ../output
    highlight: tango
    theme: cosmo
    keep_md: yes
---

```{r project_setup, include=FALSE, results='hide'}
knitr::opts_chunk$set(echo = TRUE)
#Packages required for the project
required_packages <- c('rstudioapi', 'fs', 'stringr', 'knitr')
#Installation of all required packages not already installed
install.packages(setdiff(required_packages, rownames(installed.packages())))
#Loading required packages
library('rstudioapi', 'fs', 'stringr', warn.conflicts = FALSE)
#Get this rmd file current absolute path 
#(from: https://stackoverflow.com/questions/47044068/get-the-path-of-current-script)
if (rstudioapi::isAvailable()) {
    wdir <- dirname(getActiveDocumentContext()$path)
}else{
    wdir <- getwd()
}
wdir <- stringr::str_trim(paste(head(strsplit(wdir, "/")[[1]], -1), collapse = "/"))
if (fs::dir_exists(wdir)){
    setwd(wdir)
    codedir <- file.path(wdir, "code", fsep = .Platform$file.sep)
    datadir <- file.path(wdir, "data")
    outputdir <- file.path(wdir, "output")
    features <- file.path(datadir, "features.csv")
    tryCatch(expr = {source(file.path(codedir, "myutils.R"))}
             , error = function(e){print(e)}
             , warning = function(w){print(w)})
    debugprint(c('Working dir is:', wdir))
    if (fs::dir_exists(codedir)==FALSE){ 
        print(paste('[PROBLEM] Unfound directory:', codedir))
    }else if(fs::dir_exists(datadir)==FALSE){
        print(paste('[PROBLEM] Unfound directory:', datadir))
    }else if(fs::dir_exists(outputdir)==FALSE){
        print(paste('[PROBLEM] Unfound directory:', outputdir))
    }else if(fs::file_exists(features)==TRUE){
        features <- read.csv(features)
        nfeatures <- nrow(features)
        features <- features[,2:4]
        colnames(features) <- c('Field', 'Type', 'Description')
        features$Field <- apply(features[, 1, drop = FALSE], 2, function(val){toupper(val)})
        features$Type <- apply(features[, 2, drop = FALSE], 2, function(val){tolower(val)})
    }
} else{
    print(paste('[PROBLEM] Unfound directory:', wdir))
}
```

# Loading 'events' dataset to get a first insight of its structure and contents
The 'events' dataset contains firewall log records having been tagged whether they were normal network traffic or actually related to cyber attacks. 
  - The file 'features.csv' was given to provide a description of the 'events' dataset fields. 
  - It was loaded into a table to which an extra column was added to keep the <b><i>str()</b></i> output from the 'events' fields.
  
```{r loading_dataset}
eventscsv <- file.path(datadir, "events_sample.csv") # File name needs changing here to use sample or full set
if(fs::file_exists(eventscsv)==TRUE){
    df.events <- read.csv(eventscsv)
    ncols <- ncol(df.events)
    nrows <- nrow(df.events)
    colnames <- toupper(names(df.events))
    cat(paste('The file contains', ncols, 'columns and', nrows, 'rows.\n'), fill=TRUE)
    if(nfeatures == 49){ 
        features = data.frame(features, Insight='') # Adding a column to keep the str() output
        for(i in 1:nfeatures) {
            features$Insight[i] <- capture.output(str(df.events[ , i]))
        }
    }else{
        print(str(df.events))
        print(summary(df.events))
    }
}else{
    print(paste('[PROBLEM] Unfound file:', datadir))
}
knitr::kable(features)
```

## First amendments and follow-up exploration of the 'events' dataset 
A first look at the above table allows to identify that the two last dataset fiels (ATTACK_CAT and LABEL) are most likely two response columns that will help to derive to derive a way or a model to predict the probability of facing a cyber attack based on the type of network traffic captured from this 'events' log.
The table also leads us to conclude that the initial following changes could later pay off when trying to derive the cyber attack prediction model.
  1. Change the port fields (SPORT and DSPORT) to be considered nominal fields as by nature a sequence would unlikely mean anything
  
```{r dataset_structure}
```
